The Talking Heads experiment

### 2015-04-29 chap5/chap5.tex
part 2 of martin haspelmath's comments

The semiotic square captures the four entities in 
a linguistic interaction: 
the {\itshape referent}, which is the object in the physical world that the 
speaker wants to communicate, the {\itshape segmented image}, which is 
the internal perception of the referent, the {\itshape meaning}, which is a 
category or combination of categories that picks out the 
referent in the present context, and 
the {\itshape utterance}, which is the word form 
or set of word forms transmitted by the speaker. In the previous 
chapters we looked at two sides of the semiotic
square. The relation between the real world and the
perceived image was studied in chapter 3 and the relation
between the perceived image and a conceptualisation that 
could act as the meaning for a language communication was 
studied in chapter 4. We now turn to the next side of 
the semiotic square: the relation between meaning and 
utterance.

@@

Immediately we observe a major difficulty. In realistic
language games, where agents cannot inspect each others' 
brain states nor transmit meanings directly, 
there is no feedback about the 
meaning of a word, only about the referent. For example, 
when a speaker says \smplenquote{wabo} and the hearer has correctly pointed 
to the referent that the speaker intended, neither the 
speaker nor the hearer can know whether they were 
using the same meaning. They can only know that they 
arrived at the same referent. When a game fails, the speaker 
can only point to the topic and the hearer then 
tries to figure out what possible meaning 
could have been applicable. The speaker cannot
communicate directly the \dblenquote{right} meaning, and very often more 
than one meaning is possible to distinguish a topic from 
other objects in the context, so that the hearer will 
not necessarily guess the meaning used by the speaker. 
I will call this the gavagai-problem, because Quine 
used this word to illustrate exactly this difficulty.\footnote{See \cite[29--30]{Quine:1960}.}
Quine evoked the problem of an anthropologist
trying to figure out what a native speaking an unknown
language might mean when he utters \smplenquote{gavagai} while pointing
to a white rabbit scurrying by.

@@

What cognitive architecture do agents need to engage in naming 
games? Clearly they need some sort of {\itshape associative memory} to\is{lexical memory}
store their individual lexicons.
Let us assume that agents 
can construct and recognise arbitrary consonant-vowel combinations, 
like \smplenquote{coba} or \smplenquote{wabidu}, for forming words, and that they 
have a repertoire of possible meanings in the form of 
categories, for example [\textsc{left}], [\textsc{dark}], [\textsc{large}], etc. 
The contents of the associative memory of a single agent can
be displayed in a table as follows:

\begin{table}
\begin{center}
\begin{tabular}{ l  l }
\lsptoprule
{\itshape meaning} & {\itshape form} \\ 
\midrule{}
{}[\textsc{dark}] & coba \\ 
{}[\textsc{large}] & wabidu \\ 
... & ... \\ 
\lspbottomrule
\end{tabular}
\end{center}
\caption{\label{tab:t-mem}Associative memory of a single agent.}
\end{table}
As an agent is acquiring his lexicon, there are going to 
be stages when he is not yet sure about the meaning of 
a certain form. So it must be possible for the agent
to store different 
meanings for the same form and different forms for the
same meaning. This can easily be done by extending the memory capacity
to cross-associate multiple items. Agents can then 
handle ambiguity (one word can have different
meanings) and synonymy (one meaning can 
be associated with many different words).

A speaker can only transmit a single choice for 
expressing the meaning.
When there are alternative words for the same meaning in 
his lexicon, he must decide which one to use and this decision
should be such that it maximises success in the game. 
To estimate this success, 
each agent should monitor for each form-meaning association
how successful it has been, which could be implemented 
by associating with every form-meaning pair 
a {\itshape score}.\is{score} The score of a form-meaning pair
is specific to an agent and based only on his own 
local interactions with other agents, in line with the 
principle that no single agent has a complete overview of 
the lexicon nor controls the others. An example of 
a lexicon with multiple associations and a score for every 
association is illustrated in \tabref{tab:mem2}.

\begin{table}
\begin{center}
\begin{tabular}{c c c c  c  c  c } \lsptoprule 
& coba & zapo & bila & pama & wabidu & limiri \\ 
\midrule {}		
{}[\textsc{dark}] & 0.3 & 0.2 & 0.1 & 0.8 & - & - \\ 
{}[\textsc{large}] & - & - & - & 0.5 & 0.3 & 0.6 \\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:mem2}Example lexicon with multiple associations.}
\end{center}
\end{table}
From this table we see that the agent prefers to use the 
word \smplenquote{pama} for [\textsc{dark}] and \smplenquote{limiri} for [\textsc{large}].

@@

The operations of speaker and hearer are summarised in 
\figref{incr-decr} assuming that speaker and hearer
use both the score table above (in reality they of 
course always have different score tables). 
The speaker collects all possible
forms for a given meaning [\textsc{dark}], chooses the one with the
highest score (\smplenquote{pama}), and transmits that form. 
The hearer collects all possible meanings for 
the transmitted form ([\textsc{dark}], [\textsc{large}]) and again chooses the one with
the highest score. If the hearer's meaning is equal to that
of the speaker's, the game succeeds. The score of 
two used associations increases and the others
decrease, implementing lateral 
inhibition. If the game fails, only the two used associations 
go down.\footnote{More or less neural realism can be introduced to 
model this associative memory. In our experiments we 
have a perfectly working memory that can store an 
association as soon as it has seen it once. This makes
theoretical investigation easier and makes it possible
to better follow the simulations and experiments.  
An example of a neural network solution to lexical
memory is discussed in \cite{Cangelosi:1996}.}

@@

When virgin agents start playing naming games, their 
associative memories are completely empty. Each agent 
needs two additional activities to get a lexicon
emerging: 
\begin{itemize}
\item 
When an agent does not have a word for a
meaning he wants to communicate, he is allowed to create
a new word (by random combination of vowels and 
consonants) and add that to his 
lexicon. Agents are assumed to have a 
shared repertoire of syllables which they can all 
produce and recognise. This happens with a certain 
probability, the {\itshape word creation rate} $w_{c}$. 
This rate reflects how \dblenquote{free} the agent feels to 
extend the lexicon. 
\item 
When an agent hears a word he has 
never heard before, he may add this new word to his 
repertoire. Again this happens with a certain
probability, the {\itshape word absorption rate} $w_{a}$. 
This rate reflects the critical attitude with which an 
agent accepts the linguistic authority of other agents. 
\end{itemize}
These rates are not critical but must of course
be positive. Experiments continue to work 
when agents always make a new word ($w_{c}=1$) 
and always absorb the word of the other ($w_{a}=1$).

@@

In the next game shown below, the speaker is 
{\bfshape a4}, the hearer {\bfshape a5} and the topic
{}[\textsc{small}]. Now the speaker is successful in creating a word, namely 
\smplenquote{di}. The hearer receives the word, does not know it,
but stores it in association with [\textsc{small}]. The game still fails. 
\begin{verbatim}
Game 29
   a4 is the speaker. a5 is the hearer. 
   a4 categorises the topic as [SMALL]
   a4 creates a new word: di
   a5 does not know di
   a4 points to the topic
   a5 categorises the topic as [SMALL]
   a5 stores di as [SMALL]
\end{verbatim}
In game 32, something similar happens. This time {\bfshape a5} creates
a new word \smplenquote{pida} for [\textsc{large}]. {\bfshape a3} does not know 
the word but stores it. 
\begin{verbatim}
Game 32
   a5 is the speaker. a3 is the hearer. 
   a5 categorises the topic as [LARGE]
   a5 creates a new word: \smplenquote{pida}
   a5 says: pida
   a3 does not know pida
   a5 points to the topic
   a3 categorises the topic as [LARGE]
   a3 stores pida as [LARGE]
\end{verbatim}
A first success occurs in game 43, when {\bfshape a5} uses again 
\smplenquote{pida} for [\textsc{large}]. {\bfshape a3} hears \smplenquote{pida}, has associated it 
in his lexicon with [\textsc{large}], and so the game succeeds. 
\begin{verbatim}
Game 43
   a5 is the speaker. a3 is the hearer. 
   a5 categorises the topic as [LARGE]
   a5 says: pida
   a3 interprets pida as [LARGE]
   a3 points to the topic 
   a5 signals OK
\end{verbatim}
It is quite tedious to go through such games by hand. 
For large populations of agents or meanings, even the 
most diligent researcher soon loses patience. Fortunately 
it is not so difficult to implement the Naming Game model on 
a computer. This makes large-scale simulations, even with 
hundreds of agents and meanings feasible, and ensures
that they have been done correctly.
All traces and graphs of games reported in this book
have been produced by computer simulations or physical 
experiments with the Talking Heads.

@@

\begin{table}
\begin{center}
\begin{tabular}{l  l  l }
\lsptoprule 
{\itshape Meaning} & {\itshape Word} & {\itshape Score} \\ \midrule
{}[\textsc{large}]& pida& 0.20\\ 
  & fobu& 0.1\\ 
{}[\textsc{light}]& gi& 0.0\\ 
{}[\textsc{small}]& di & 0.10\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-mem3}Associative memory of a single agent after 100 games.}
\end{center}
\end{table}
Both the words \smplenquote{di} and \smplenquote{pida} are present but with 
weak scores. There are two synonyms for 
{}[\textsc{large}]: \smplenquote{pida} and \smplenquote{fobu}, but \smplenquote{pida} is 
preferred. There is a word \smplenquote{gi} who is available 
for [\textsc{light}] but does not have a positive score,
because successive trials failed to yield a successful game.

A table such as the one above only represents the lexicon
of a single agent. It is highly unlikely that two agents
share the same lexicon because each agent will have 
had different encounters and hence different language
experiences. A picture of the lexicon of the group from the 
viewpoint of an outside observer can be obtained by
inspecting the internal states of each agent to 
construct the {\itshape group lexicon}. It groups the 
dominating meaning-form associations for all possible
meanings and the frequency of each association. 
It gives a picture of \dblenquote{the} lexicon 
in the group. The group lexicon
for the complete population of five agents after 50 language games
is shown in \tabref{tab:t-mem4}.

\begin{table}
\begin{center}
\begin{tabular}{l  l  l }
\lsptoprule 
{\itshape Meaning}& {\itshape Word} & {\itshape Frequency} \\ \midrule 
{}[\textsc{large}]& pida& 0.40\\ 
{}[\textsc{light}]& gi & 0.60\\ 
{}[\textsc{small}]& di & 0.40\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-mem4}Population lexicon after 50 games}
\end{center}
\end{table}
This reflects a situation where 40 \%
of the agents prefers to name [\textsc{large}] with
\smplenquote{pida}. 60 \% of the agents
use \smplenquote{gi} for [\textsc{light}], and 40 \% \smplenquote{di} for [\textsc{small}]. 
The other meanings ([\textsc{dark}] and [\textsc{red}]) do not have names yet.

Note that the group lexicon is not known by the agents
and is not stored anywhere in the 
total system. The only information which is locally stored in each 
agent is his own lexicon, which might be quite different from that 
of the group lexicon. For example, the lexicon of {\bfshape a5} shown earlier
is different from the group lexicon.
{\bfshape a5}'s score for \smplenquote{gi} is 0.0
even though the word is already preferred by 60 \% of 
the agents according to the group lexicon. 
The group lexicon is a macroscopic structure that we
as observers construct from inspecting the internal 
states of the agents.

Let us now continue the simulation. Here are two additional
games showing how \smplenquote{ga} propagates from {\bfshape a4} to {\bfshape a2} in 
game 101 and from {\bfshape a2} to {\bfshape a1} in game 104. 
\begin{verbatim}
Game 104
   a4 is the speaker. a2 is the hearer. 
   a4 categorises the topic as [RED]
   a4 says: ga
   a2 does not know ga
   a4 points to the topic
   a2 categorises the topic as [RED]
   a2 stores ga as [RED]
\end{verbatim}
\begin{verbatim}
Game 104
   a2 is the speaker. a1 is the hearer. 
   a2 categorises the topic as [RED]
   a1 says: ga
   a1 does not know ga
   a2 points to the topic
   a1 categorises the topic as [RED]
   a1 stores ga as [RED]
\end{verbatim}
After a total of 250 games, the consensus is complete. 
The group lexicon is shown in \tabref{tab:t-mem5}.

@@

I use various measures both for actual lexicon use and for 
the coherence and evolution of the lexicons of
the individuals to see better what is happening. 
The first and easiest measure is the {\itshape average game success}, 
also called the {\itshape communicative success}, \is{communicative success}
of a population of agents ${\cal A}$ in a set of $n$ language games. 
When this measure is graphed continuously for consecutive sets of 
games, the progress in the population
towards successful communication can be followed easily. This
is shown in \figref{success} which plots data from the 
games discussed in the previous paragraphs. We see at once that
average success climbs from a starting point of zero
to a maximum of 1.0. This can only be because a shared lexicon 
emerged in the population.

@@

To monitor to what extent the agents share the same
lexicon, I propose a second measure, the {\itshape lexical
coherence}. The lexical coherence is 
defined as the average of the frequencies of
all the form-meaning pairs in the group lexicon.
If all agents prefer the same form-meaning 
pair for all meanings, lexical coherence is 1.0. If they agree
on none, it is 0.0.\is{lexical coherence}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap5/figs/suc-coh.pdf}}
\caption{\label{suc-coh}This figure shows the evolution of both average
success and lexical coherence for a group of 20 agents and 20 
meanings. Total lexical coherence climbs more slowly
once the 
population has reached a high average success.}
\end{figure}
Does total success imply that all agents use the 
same lexicon? Not really. To have success, the hearer
must associate the form used by the speaker
with the same meaning. But it is not required that the
hearer himself prefers to use the same form for the same meaning, 
synonyms may occur.  
A speaker of British English typically uses the word \smplenquote{pavement}, 
whereas an American prefers \smplenquote{sidewalk}, even though he
understands \smplenquote{pavement}. Thus there can 
be several forms active in the same population, even though 
the outcome of a game is always successful. We will see 
later that synonyms do get damped, as is the case in 
human natural lexicons.

@@

The most remarkable and at first mysterious property of the 
Naming Game model is that the agents somehow reach a consensus
without any central supervisor. They do {\itshape not} do this by 
having a general overview or by changing their internal 
parameters so as to become more conservative as the lexicon
solidifies. It is solely due to the subtle interaction between
language use, which gradually becomes uniform, and each agent's
adaptation to the language heard in the environment. 
If a certain word comes to be preferred by a group of agents
for a certain meaning, its frequency of use goes up so that
others encounter this 
word more often and hence their scores for that word
continue to increase as well.
The more agents use a word, the higher its chance of success
and the more it will be used. This effect is still enforced 
by lateral inhibition. The scores of competing associations 
decrease, making it less likely that they will win in the future. 
This positive feedback therefore introduces an\is{autocatalysis} 
{\itshape autocatalytic} (self-enforcing) effect until the population
locks into an equilibrium state.\footnote{
Such positive feedback loops and the stability criteria
associated with them have been widely studied in non-linear
dynamical systems and applied to chemical and biological 
processes. See \cite{Babloyantz:1986}.}

To follow better how a consensus gradually emerges, I will visualise
the competition between different words for the same meaning
in a {\itshape meaning-form (MF) competition diagram}, 
such as the one in \figref{competition}, which monitors 
the frequencies of the different forms in use for one meaning.
The diagram shows clearly the struggle between different forms until
one form (\smplenquote{pe}) emerges as the winner. When we later study 
grounded lexicon formation processes, we will see that the 
competition becomes much more complex and the whole system 
is in constant evolution. A form-meaning pair which is 
dominating may become weaker because its meaning
fails to pick up the right
referent in a new context. 
This in turn may trigger the creation of new words or the 
resurgence of existing words.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap5/figs/comp1.pdf}}
\caption{\label{competition}Simulation with 
a population of 20 agents. The meaning-form competition 
diagram shows the frequency of all competing forms for a 
single meaning. We see a winner-take-all situation
with one word (\smplenquote{pe}) dominating.}
\end{figure}

@@

\subsection{Spatially Distributed Naming Games}

@@

Initially we let each subgroup evolve towards a shared 
communication. 
Success is never total because there are occasional interactions
with members of other communities, however inner-cluster
communication reaches total success.
However inspection of the agent lexicons reveals that
agents will develop a stable
language within their cluster, but also a
second language, an {\itshape interlingua}, which is weaker but
shared among the different clusters. This interlingua will
become stronger as more agents interact between clusters.
Thus we observe language diversity due to
the spatial distribution but at the same time the rise 
of an interlingua.

>>>>

The semiotic square captures the four entities in 
a linguistic interaction: 
the {\scshape referent}, which is the object in the physical world that the 
speaker wants to communicate, the {\scshape segmented image}, which is 
the internal perception of the referent, the {\scshape meaning}, which is a 
category or combination of categories that picks out the 
referent in the present context, and 
the {\scshape utterance}, which is the word form 
or set of word forms transmitted by the speaker. In the previous 
chapters we looked at two sides of the semiotic
square. The relation between the real world and the
perceived image was studied in chapter 3 and the relation
between the perceived image and a conceptualisation that 
could act as the meaning for a language communication was 
studied in chapter 4. We now turn to the next side of 
the semiotic square: the relation between meaning and 
utterance.

@@

Immediately we observe a major difficulty. In realistic
language games, where agents cannot inspect each others' 
brain states nor transmit meanings directly, 
there is no feedback about the 
meaning of a word, only about the referent. For example, 
when a speaker says \textit{wabo} and the hearer has correctly pointed 
to the referent that the speaker intended, neither the 
speaker nor the hearer can know whether they were 
using the same meaning. They can only know that they 
arrived at the same referent. When a game fails, the speaker 
can only point to the topic and the hearer then 
tries to figure out what possible meaning 
could have been applicable. The speaker cannot
communicate directly the \dblenquote{right} meaning, and very often more 
than one meaning is possible to distinguish a topic from 
other objects in the context, so that the hearer will 
not necessarily guess the meaning used by the speaker. 
I will call this the gavagai-problem, because Quine 
used this word to illustrate exactly this difficulty.\footnote{See \cite[29--30]{Quine:1960}.}
Quine evoked the problem of an anthropologist
trying to figure out what a native speaking an unknown
language might mean when he utters \textit{gavagai} while pointing
to a white rabbit scurrying by.

@@

What cognitive architecture do agents need to engage in naming 
games? Clearly they need some sort of {\scshape associative memory} to\is{lexical memory}
store their individual lexicons.
Let us assume that agents 
can construct and recognise arbitrary consonant-vowel combinations, 
like \textit{coba} or \textit{wabidu}, for forming words, and that they 
have a repertoire of possible meanings in the form of 
categories, for example [\textsc{left}], [\textsc{dark}], [\textsc{large}], etc. 
The contents of the associative memory of a single agent can
be displayed in a table as follows:

\begin{table}
\begin{center}
\begin{tabular}{ l  l }
\lsptoprule
{\itshape meaning} & {\itshape form} \\ 
\midrule{}
{}[\textsc{dark}] & coba \\ 
{}[\textsc{large}] & wabidu \\ 
... & ... \\ 
\lspbottomrule
\end{tabular}
\end{center}
\caption{\label{tab:t-mem}Associative memory of a single agent}
\end{table}
As an agent is acquiring his lexicon, there are going to 
be stages when he is not yet sure about the meaning of 
a certain form. So it must be possible for the agent
to store different 
meanings for the same form and different forms for the
same meaning. This can easily be done by extending the memory capacity
to cross-associate multiple items. Agents can then 
handle ambiguity (one word can have different
meanings) and synonymy (one meaning can 
be associated with many different words).

A speaker can only transmit a single choice for 
expressing the meaning.
When there are alternative words for the same meaning in 
his lexicon, he must decide which one to use and this decision
should be such that it maximises success in the game. 
To estimate this success, 
each agent should monitor for each form-meaning association
how successful it has been, which could be implemented 
by associating with every form-meaning pair 
a {\scshape score}.\is{score} The score of a form-meaning pair
is specific to an agent and based only on his own 
local interactions with other agents, in line with the 
principle that no single agent has a complete overview of 
the lexicon nor controls the others. An example of 
a lexicon with multiple associations and a score for every 
association is illustrated in \tabref{tab:mem2}.

\begin{table}
\begin{center}
\begin{tabular}{c c c c  c  c  c } \lsptoprule 
& \textit{coba} & \textit{zapo} & \textit{bila} & \textit{pama} & \textit{wabidu} & \textit{limiri} \\ 
\midrule {}		
{}[\textsc{dark}] & 0.3 & 0.2 & 0.1 & 0.8 & - & - \\ 
{}[\textsc{large}] & - & - & - & 0.5 & 0.3 & 0.6 \\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:mem2}Example lexicon with multiple associations}
\end{center}
\end{table}
From this table we see that the agent prefers to use the 
word \textit{pama} for [\textsc{dark}] and \textit{limiri} for [\textsc{large}].

@@

The operations of speaker and hearer are summarised in 
\figref{incr-decr} assuming that speaker and hearer
use both the score table above (in reality they of 
course always have different score tables). 
The speaker collects all possible
forms for a given meaning [\textsc{dark}], chooses the one with the
highest score (\textit{pama}), and transmits that form. 
The hearer collects all possible meanings for 
the transmitted form ([\textsc{dark}], [\textsc{large}]) and again chooses the one with
the highest score. If the hearer's meaning is equal to that
of the speaker's, the game succeeds. The score of 
two used associations increases and the others
decrease, implementing lateral 
inhibition. If the game fails, only the two used associations 
go down.\footnote{More or less neural realism can be introduced to 
model this associative memory. In our experiments we 
have a perfectly working memory that can store an 
association as soon as it has seen it once. This makes
theoretical investigation easier and makes it possible
to better follow the simulations and experiments.  
An example of a neural network solution to lexical
memory is discussed in \cite{Cangelosi:1996}.}

@@

When virgin agents start playing naming games, their 
associative memories are completely empty. Each agent 
needs two additional activities to get a lexicon
emerging: 
\begin{itemize}
\item 
When an agent does not have a word for a
meaning he wants to communicate, he is allowed to create
a new word (by random combination of vowels and 
consonants) and add that to his 
lexicon. Agents are assumed to have a 
shared repertoire of syllables which they can all 
produce and recognise. This happens with a certain 
probability, the {\scshape word creation rate} $w_{c}$. 
This rate reflects how \dblenquote{free} the agent feels to 
extend the lexicon. 
\item 
When an agent hears a word he has 
never heard before, he may add this new word to his 
repertoire. Again this happens with a certain
probability, the {\scshape word absorption rate} $w_{a}$. 
This rate reflects the critical attitude with which an 
agent accepts the linguistic authority of other agents. 
\end{itemize}
These rates are not critical but must of course
be positive. Experiments continue to work 
when agents always make a new word ($w_{c}=1$) 
and always absorb the word of the other ($w_{a}=1$).

@@

In the next game shown below, the speaker is 
{\bfshape a4}, the hearer {\bfshape a5} and the topic
{}[\textsc{small}]. Now the speaker is successful in creating a word, namely 
\textit{di}. The hearer receives the word, does not know it,
but stores it in association with [\textsc{small}]. The game still fails. 
\begin{verbatim}
Game 29
   a4 is the speaker. a5 is the hearer. 
   a4 categorises the topic as [SMALL]
   a4 creates a new word: di
   a5 does not know di
   a4 points to the topic
   a5 categorises the topic as [SMALL]
   a5 stores di as [SMALL]
\end{verbatim}
In game 32, something similar happens. This time {\bfshape a5} creates
a new word \textit{pida} for [\textsc{large}]. {\bfshape a3} does not know 
the word but stores it. 
\begin{verbatim}
Game 32
   a5 is the speaker. a3 is the hearer. 
   a5 categorises the topic as [LARGE]
   a5 creates a new word: pida
   a5 says: pida
   a3 does not know pida
   a5 points to the topic
   a3 categorises the topic as [LARGE]
   a3 stores pida as [LARGE]
\end{verbatim}
A first success occurs in game 43, when {\bfshape a5} uses again 
\textit{pida} for [\textsc{large}]. {\bfshape a3} hears \textit{pida}, has associated it 
in his lexicon with [\textsc{large}], and so the game succeeds. 
\begin{verbatim}
Game 43
   a5 is the speaker. a3 is the hearer. 
   a5 categorises the topic as [LARGE]
   a5 says: pida
   a3 interprets pida as [LARGE]
   a3 points to the topic 
   a5 signals OK
\end{verbatim}
It is quite tedious to go through such games by hand. 
For large populations of agents or meanings, even the 
most diligent researcher soon loses patience. Fortunately 
it is not so difficult to implement the Naming Game model on 
a computer. This makes large-scale simulations, even with 
hundreds of agents and meanings feasible, and ensures
that they have been done correctly.
All traces and graphs of games reported in this book
have been produced by computer simulations or physical 
experiments with the Talking Heads.

@@

\begin{table}
\begin{center}
\begin{tabular}{l  l  l }
\lsptoprule 
{\itshape Meaning} & {\itshape Word} & {\itshape Score} \\ \midrule
{}[\textsc{large}]& pida& 0.20\\ 
  & fobu& 0.1\\ 
{}[\textsc{light}]& gi& 0.0\\ 
{}[\textsc{small}]& di & 0.10\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-mem3}Associative memory of a single agent after 100 games.}
\end{center}
\end{table}
Both the words \textit{di} and \textit{pida} are present but with 
weak scores. There are two synonyms for 
{}[\textsc{large}]: \textit{pida} and \textit{fobu}, but \textit{pida} is 
preferred. There is a word \textit{gi} who is available 
for [\textsc{light}] but does not have a positive score,
because successive trials failed to yield a successful game.

A table such as the one above only represents the lexicon
of a single agent. It is highly unlikely that two agents
share the same lexicon because each agent will have 
had different encounters and hence different language
experiences. A picture of the lexicon of the group from the 
viewpoint of an outside observer can be obtained by
inspecting the internal states of each agent to 
construct the {\scshape group lexicon}. It groups the 
dominating meaning-form associations for all possible
meanings and the frequency of each association. 
It gives a picture of \dblenquote{the} lexicon 
in the group. The group lexicon
for the complete population of five agents after 50 language games
is shown in \tabref{tab:t-mem4}.

\begin{table}
\begin{center}
\begin{tabular}{l  l  l }
\lsptoprule 
{\itshape Meaning}& {\itshape Word} & {\itshape Frequency} \\ \midrule 
{}[\textsc{large}]& pida& 0.40\\ 
{}[\textsc{light}]& gi & 0.60\\ 
{}[\textsc{small}]& di & 0.40\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-mem4}Population lexicon after 50 games}
\end{center}
\end{table}
This reflects a situation where 40 \%
of the agents prefers to name [\textsc{large}] with
\textit{pida}. 60 \% of the agents
use \textit{gi} for [\textsc{light}], and 40 \% \textit{di} for [\textsc{small}]. 
The other meanings ([\textsc{dark}] and [\textsc{red}]) do not have names yet.

Note that the group lexicon is not known by the agents
and is not stored anywhere in the 
total system. The only information which is locally stored in each 
agent is his own lexicon, which might be quite different from that 
of the group lexicon. For example, the lexicon of {\bfshape a5} shown earlier
is different from the group lexicon.
{\bfshape a5}'s score for \textit{gi} is 0.0
even though the word is already preferred by 60 \% of 
the agents according to the group lexicon. 
The group lexicon is a macroscopic structure that we
as observers construct from inspecting the internal 
states of the agents.

Let us now continue the simulation. Here are two additional
games showing how \textit{ga} propagates from {\bfshape a4} to {\bfshape a2} in 
game 101 and from {\bfshape a2} to {\bfshape a1} in game 104. 
\begin{verbatim}
Game 104
   a4 is the speaker. a2 is the hearer. 
   a4 categorises the topic as [RED]
   a4 says: ga
   a2 does not know ga
   a4 points to the topic
   a2 categorises the topic as [RED]
   a2 stores ga as [RED]
\end{verbatim}
\begin{verbatim}
Game 104
   a2 is the speaker. a1 is the hearer. 
   a2 categorises the topic as [RED]
   a1 says: ga
   a1 does not know ga
   a2 points to the topic
   a1 categorises the topic as [RED]
   a1 stores ga as [RED]
\end{verbatim}
After a total of 250 games, the consensus is complete. 
The group lexicon is shown in \tabref{tab:t-mem5}.

@@

I use various measures both for actual lexicon use and for 
the coherence and evolution of the lexicons of
the individuals to see better what is happening. 
The first and easiest measure is the {\scshape average game success}, 
also called the {\scshape communicative success}, \is{communicative success}
of a population of agents ${\cal A}$ in a set of $n$ language games. 
When this measure is graphed continuously for consecutive sets of 
games, the progress in the population
towards successful communication can be followed easily. This
is shown in \figref{success} which plots data from the 
games discussed in the previous paragraphs. We see at once that
average success climbs from a starting point of zero
to a maximum of 1.0. This can only be because a shared lexicon 
emerged in the population.

@@

To monitor to what extent the agents share the same
lexicon, I propose a second measure, the {\scshape lexical
coherence}. The lexical coherence is 
defined as the average of the frequencies of
all the form-meaning pairs in the group lexicon.
If all agents prefer the same form-meaning 
pair for all meanings, lexical coherence is 1.0. If they agree
on none, it is 0.0.\is{lexical coherence}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap5/figs/suc-coh.pdf}}
\caption{\label{suc-coh}This figure shows the evolution of both average
success and lexical coherence for a group of 20 agents and 20 
meanings. Total lexical coherence climbs more slowly
once the 
population has reached a high average success.}
\end{figure}
Does total success imply that all agents use the 
same lexicon? Not really. To have success, the hearer
must associate the form used by the speaker
with the same meaning. But it is not required that the
hearer himself prefers to use the same form for the same meaning, 
synonyms may occur.  
A speaker of British English typically uses the word \textit{pavement}, 
whereas an American prefers \textit{sidewalk}, even though he
understands \textit{pavement}. Thus there can 
be several forms active in the same population, even though 
the outcome of a game is always successful. We will see 
later that synonyms do get damped, as is the case in 
human natural lexicons.

@@

The most remarkable and at first mysterious property of the 
Naming Game model is that the agents somehow reach a consensus
without any central supervisor. They do {\itshape not} do this by 
having a general overview or by changing their internal 
parameters so as to become more conservative as the lexicon
solidifies. It is solely due to the subtle interaction between
language use, which gradually becomes uniform, and each agent's
adaptation to the language heard in the environment. 
If a certain word comes to be preferred by a group of agents
for a certain meaning, its frequency of use goes up so that
others encounter this 
word more often and hence their scores for that word
continue to increase as well.
The more agents use a word, the higher its chance of success
and the more it will be used. This effect is still enforced 
by lateral inhibition. The scores of competing associations 
decrease, making it less likely that they will win in the future. 
This positive feedback therefore introduces an\is{autocatalysis} 
{\scshape autocatalytic} (self-enforcing) effect until the population
locks into an equilibrium state.\footnote{
Such positive feedback loops and the stability criteria
associated with them have been widely studied in non-linear
dynamical systems and applied to chemical and biological 
processes. See \cite{Babloyantz:1986}.}

To follow better how a consensus gradually emerges, I will visualise
the competition between different words for the same meaning
in a {\scshape meaning-form (MF) competition diagram}, 
such as the one in \figref{competition}, which monitors 
the frequencies of the different forms in use for one meaning.
The diagram shows clearly the struggle between different forms until
one form (\textit{pe}) emerges as the winner. When we later study 
grounded lexicon formation processes, we will see that the 
competition becomes much more complex and the whole system 
is in constant evolution. A form-meaning pair which is 
dominating may become weaker because its meaning
fails to pick up the right
referent in a new context. 
This in turn may trigger the creation of new words or the 
resurgence of existing words.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap5/figs/comp1.pdf}}
\caption{\label{competition}Simulation with 
a population of 20 agents. The meaning-form competition 
diagram shows the frequency of all competing forms for a 
single meaning. We see a winner-take-all situation
with one word (\textit{pe}) dominating.}
\end{figure}

@@

\subsection{Spatially distributed naming games}

@@

Initially we let each subgroup evolve towards a shared 
communication. 
Success is never total because there are occasional interactions
with members of other communities, however inner-cluster
communication reaches total success.
However inspection of the agent lexicons reveals that
agents will develop a stable
language within their cluster, but also a
second language, an {\scshape interlingua}, which is weaker but
shared among the different clusters. This interlingua will
become stronger as more agents interact between clusters.
Thus we observe language diversity due to
the spatial distribution but at the same time the rise 
of an interlingua.
